---
title: "Reproducible Research: Peer Assessment 1"
author: "Ivana Seric"
date: "August 9, 2015"
output: 
html_document:
keep_md: true
---


This is the first project for "Repsoducibe Research" course, fifth course in the Coursera "Data Science" specialization. 

```{r setup, results='hide', message=FALSE, warning=FALSE}
setwd('~/R/Reproducible_research/RepData_PeerAssessment1/')
library(plyr)
library(lubridate)
library(lattice)
options(scipen = 1, digits = 2)
```

## Loading and preprocessing the data

1. Load the data

```{r reading data, cache = TRUE}
data_download <- mdy("August 9, 2015")
myData <- read.csv("activity.csv")
```

The data was downloaded from the coarse website on `r data_download`.

2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r processing data}
# change date factor into "POSIXct", for easier date maniplation
myData$date <- ymd(as.character(myData$date))
```


## What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

1. Calculate the total number of steps taken per day
```{r total steps}
totalsByDay <- ddply(myData, "date", numcolwise(sum))$steps
```

2. Make a histogram of the total number of steps taken each day 
```{r histogram, fig.height=4, fig.width=6}
hist(totalsByDay, col = "green", ann = FALSE, breaks = 10)
title(xlab = "Number of steps", ylab = "Frequency", main = "Total number of steps per day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r mean and median, results="asis"}
stepsMean <- mean(totalsByDay, na.rm = TRUE)
stepsMedian <- median(totalsByDay, na.rm = TRUE)
```

The mean of the total number of steps taken per day is `r stepsMean`, 
and the median is `r stepsMedian`

## What is the average daily activity pattern?


1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r average by interval, fig.height=4, fig.width=6}
sumByInterval <- ddply(myData, "interval", na.rm = TRUE, numcolwise(sum))
numOfdays <- length(totalsByDay)
avStepsInterval <- sumByInterval$steps/numOfdays
plot(sumByInterval$interval, avStepsInterval, type = "l", col = "blue", lwd = 2, ann = FALSE)
title(xlab = "Interval", ylab = "Av. Steps", main = "Average steps for intervals")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?


```{r maximum activty interval}
maxIndex <- which(max(sumByInterval$steps, na.rm = TRUE) == sumByInterval$steps, arr.ind = TRUE)
maxInterval <- sumByInterval$interval[maxIndex]
timeHour <- maxInterval %/% 100
timeMinute <- maxInterval %% 100
```

Time interval  `r timeHour`:`r timeMinute` - `r timeHour`:`r timeMinute+5` contains the maximum number of steps.


## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r count NAs}
numNAs <- sum(is.na(myData$steps))
```

The total number of missing values is `r numNAs`.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.  
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r filling NAs }
# Replace NAs with the mean for that interval
dataLength <- dim(myData)[1]
noNAs <- rep(0, dataLength)

for (i in 1:dataLength) {
    if (is.na(myData$steps[i])) {
        intervalIndex <- which(sumByInterval$interval == myData$interval[i], arr.ind = TRUE)
        noNAs[i] <- avStepsInterval[intervalIndex]
    } else {
        noNAs[i] <- myData$steps[i]
    }
}

myDataNew <- data.frame(noNAs, myData$date, myData$interval)
colnames(myDataNew) <- c("steps", "date", "interval")
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r new historgam, fig.height=4, fig.width=6}
totalsByDayNew <- ddply(myDataNew, "date", numcolwise(sum))$steps
hist(totalsByDayNew, col = "green", ann = FALSE, breaks = 10)
title(xlab = "Number of steps", ylab = "Frequency", main = "Total number of steps per day without NAs")
```

```{r mean and median new, }
stepsMeanNew <- mean(totalsByDayNew)
stepsMedianNew <- median(totalsByDayNew)
```

The new mean of the total number of steps taken per day is `r stepsMeanNew`, 
and the new median is `r stepsMedianNew`. 


## Are there differences in activity patterns between weekdays and weekends?

For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.

1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r add weekday/end factor}
dataWeekday <- weekdays(myData$date, abbreviate = TRUE)
weekdayFactor <- as.factor( ifelse(dataWeekday == c("Sun", "Sat"), "weekend", "weekday" ))
myDataNew <- cbind(myDataNew, weekdayFactor)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r panel plot}
sumByIntervalNew <- ddply(myDataNew, .(interval, weekdayFactor), numcolwise(sum))
wdays <- data.frame(table(myDataNew$weekdayFactor))
totalWeekends <- wdays$Freq[which(wdays$Var1 == "weekend")]
totalWeekdays <- wdays$Freq[which(wdays$Var1 == "weekday")]
totalDaysPerFactor <- ifelse(sumByIntervalNew$weekdayFactor == "weekend", totalWeekends, totalWeekdays )
avStepsIntervalNew <- sumByIntervalNew$steps/totalDaysPerFactor

xyplot(avStepsIntervalNew ~ sumByIntervalNew$interval | sumByIntervalNew$weekdayFactor, type = "l", col = "blue", lwd = 2, main = "Average steps for intervals", xlab = "interval", ylab = "average steps")

```






